---
title: "Vectorisation"
date: "`r Sys.Date()`"
output:
  rmdformats::html_clean:
    highlight: kate
    code_folding: hide
---


```{r setup, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

# Intro

To convert this notebook into HTML (to *knit* it), type `make` in the command line of this folder or click **Knit** icon in the top bar of the RStudio editor.

R is a *high-level, interpreted* computer language. This means that R takes care of a lot of basic computer tasks for you.

Most of R’s functions are vectorized, meaning that the function will operate on all elements of a vector without needing to loop through and act on each element one at a time. This makes writing code more concise, easy to read, and less error prone.

# Adding two vectors in a loop

Suppose we need to add 10M numbers to each other. This would be a *classical* way of doing it using a for loop:

```{r, echo = T}
nNum = 1e7
vA = rnorm(nNum)
vB = rnorm(nNum)
vCloop = rep(0., nNum)

system.time(
  for(ii in 1:nNum) {
    vCloop[ii] = vA[ii] + vB[ii]
  }  
)
```

Instead, R provides a convenient operator `+` that also works with vectors!

```{r, echo = T}
system.time({vCvec = vA + vB})
```

The result of two operations is exactly the same, but the latter is way faster!

```{r, echo = T}
all.equal(vCloop, vCvec)
```

```{r, echo=FALSE}
# clean
rm(nNum,
   vA, vB, vCloop, vCvec)
```

# Inefficient for loop

**However**, for loops can be fast **if** you follow some rules:

1. Don’t use a loop when a vectorized alternative exists (see the previous slide).
2. Don't increase the size of your objects, e.g. using `cbind`, `rbind`, during the loop! Instead, pre-allocate memory by predefining an object to hold the result. 

Example of an inefficient loop: at every iteration of the loop, `myRes` increases in size because a new element is added to this vector.

```{r, echo = T}
nNum = 1e4
vA = rnorm(nNum)
vB = rnorm(nNum)

myRes = c()
system.time(
  for(ii in 1:nNum) {
    myRes = c(myRes, vA[ii] + vB[ii])
  }  
)
```

# Efficient for loop

```{r, echo = T}
# Pre-allocated vector where the result will be written to
vC = rep(0., nNum)

system.time(
  for(ii in seq_along(vA)) {
    vC[ii] = vA[ii] + vB[ii]
  }  
)
```

```{r, echo = T}
all.equal(myRes, vC)
```

The result is the same, but pre-allocating the `vC` vector speeds up the computation by two orders of magnitude!

The difference between the computation time increases with the size of the vector because without pre-allocating the memory, R is copying the existing vector before increasing its size with the `c` operation.

```{r, echo = F}
# Clean
rm(nNum,
   vA, vB, myRes, vC)
```

